---
title: "NYPD Shooting Incident Data (Historic)"
author: "J Cropsey"
date: '2022-05-17'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

## NYPD Shooting Incident Data (Historic)

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of 2020.

```{r readData}
NYPD <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
summary(NYPD)
```

## Tidying data
Removing unnecessary columns, and changing data types of multiple columns.

Many incidents are missing certain data points. Depending on the analysis required, we may need to filter them out of the data. These can be easily identified with the ```is.na()``` function.

```{r tidy}
NYPD <- NYPD %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>% 
  mutate(PERP_AGE_GROUP = factor(PERP_AGE_GROUP)) %>% 
  mutate(PERP_SEX = factor(PERP_SEX))  %>% 
  mutate(PERP_RACE = factor(PERP_RACE))  %>% 
  mutate(VIC_AGE_GROUP = factor(VIC_AGE_GROUP)) %>% 
  mutate(VIC_SEX = factor(VIC_SEX)) %>% 
  mutate(VIC_RACE = factor(VIC_RACE)) %>% 
  mutate(STATISTICAL_MURDER_FLAG = factor(STATISTICAL_MURDER_FLAG)) %>% 
  mutate(BORO = factor(BORO)) %>% 
  mutate(LOCATION_DESC = factor(LOCATION_DESC)) %>% 
  mutate(PRECINCT = factor(PRECINCT)) %>% 
  mutate(JURISDICTION_CODE = factor(JURISDICTION_CODE)) %>% 
  mutate(Lon_Lat = factor(Lon_Lat)) %>%
  separate(Lon_Lat, sep = " ", into = c("Point","Lon_Lat", "Test")) %>%
  unite(Lon_Lat, Test, col = "Lon_Lat", sep = ", ") %>%
  mutate(Lon_Lat = factor(Lon_Lat))
NYPD <- select(NYPD, -c(Point, Latitude, Longitude, X_COORD_CD, Y_COORD_CD, INCIDENT_KEY))

summary(NYPD)
```

## Visualizations and Analysis

First I was curious where most of the shooting incidents were taking place. I thought it would be good to break it down by Boro.

```{r locations}
NYPD %>%
  mutate(BORO = BORO %>% fct_infreq() %>% fct_rev()) %>% ggplot(aes(BORO)) +
  geom_bar()
```

Fifteen years is a long time, so I wanted to look at a slightly smaller sample. And I was curious specifically about unsolved murders. Incidents that were missing information on the perpetrator were taken to be unsolved.

```{r five_years}
unsolved_five_years <- NYPD %>%
  filter(is.na(PERP_SEX), STATISTICAL_MURDER_FLAG == TRUE, year(OCCUR_DATE) > 2015) 
  unsolved_five_years %>%
  mutate(BORO = BORO %>% fct_infreq() %>% fct_rev()) %>% ggplot(aes(BORO)) +
  geom_bar()
```

It's interesting that unsolved murders in the last five years, and total shooting incidents in the last fifteen years, have very similar distributions over the NYC Boros.

I then wanted to see how the number of shooting incidents has changed over time.
```{r histo}
NYPD %>%
     ggplot(aes(x = OCCUR_DATE)) +
     geom_histogram(bins = 60)
```

Then I wanted to try and model for the number of murders in NYC, based on shooting incidents. I created two new variables, incidents per year and murders per year, and then ran a linear model to see if incidents per year predicted murders per year.
```{r model}
NYPD_per_year <- NYPD %>%
  add_count(year(OCCUR_DATE), name = "INCIDENTS_IN_YR") %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  add_count(year(OCCUR_DATE), name = "MURDERS_IN_YR")


mod <- lm(MURDERS_IN_YR ~ INCIDENTS_IN_YR, data = NYPD_per_year)
summary(mod)

NYPD_per_year <- NYPD_per_year %>%
  mutate(pred = predict(mod))

NYPD_per_year %>% ggplot() +geom_point(aes(x = INCIDENTS_IN_YR, y = MURDERS_IN_YR), color = "blue") + 
  geom_point(aes(x = INCIDENTS_IN_YR, y = pred), color = "red")
```   
     
The model does a decent job of predicting the number of murders. Clearly, the more shooting incidents in the city, the more murders occur. There are clearly some other variables that would need to be added to make the model even more accurate though. 

Based on all of this analysis, it looks like shooting incidents have had some peaks and valleys, but for the most part have stayed mostly flat over the last 15 years. It also appears that Brooklyn is the most dangerous of the five NYC Boros.

This project does have some possible sources of bias though. For one thing, it's unclear how these incidents were reported. It's possible that Brooklyn does the best job of reporting shooting incidents, making it look more dangerous than it is. Also, the data did not include population data, so it is possible that the distribution of incidents in simply related to the population of each Boro.
